A Formal Theory of Bidirectional Typing for Simple Types

### 2023-05-25 tex/BiSig.tex
[ update ] Fine-tune typing rules

\documentclass[acmsmall]{acmart}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[british]{babel}

@@

\subsubsection{Algebraic theory of type theories}
\cite{Fiore1999,Fiore2022,Hamana2022,Ahrens2021,Arkor2020}

@@

More importantly, the inclusion of the set of variables used in an operation is a salient feature of our binding signatures.
Instead of treating the application construct $\app_{A, B}$ as indexed by all types $A$ and $B$, as done by \citet{Fiore2022}, we are able to identify them as a single construct $\app$.
This is not only for brevity but also necessary to incorporate the global context of type variables available, which is crucial for unification to solve constraints during type synthesis.

In contrast to the set $\Theta$ representing the context of global type variables for a derivation of a typing judgement $\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A$ introduced below, the set $\Xi$ associated with an operation can be referred to as the \emph{local context} for that operation.

@@

The substitution $\rho\colon \Sub{\Xi}{\Theta}$ is used to instantiate variables in the local context $\Xi$ with global type variables in $\Theta$.
Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\termOp_o$ to ensure all types are will-formed under $\Theta$.

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A} \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$ and $\Theta \vdash_{\Sigma} A$} \\
    \inferrule{A \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
    \and
    \inferrule*{\rho : \Sub{\Xi}{\Theta} \\ \Theta \mid \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\termOp)
  \end{mathpar}
  for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
  \caption{Intrinsic typing rules}
  \label{fig:intrinsic-typing}
\end{figure}
\subsection{Raw Terms and Extrinsically Typed Terms?}
\begin{LTannote}{no extrinsic typing}
Argue that we don't need extrinsically typed terms but raw terms only as the output of some parser.
\end{LTannote}

@@

\begin{definition}
  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Cxt_{\Sigma}(\Xi) \times \Mode \times \Type_{\Sigma}(\Xi)\right)^* \times \Mode \times \Type_{\Sigma}(\Xi).
  \]
  where $\Mode$ consists of $\chk$ for checking and $\syn$ for synthesis.
  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\termOp_o$ in a bidirectional system, i.e.
  an arity is a $4$-tuple
  \[
    \left(\Xi, \left[\left(\Delta_1, d_1, A_1\right), \ldots, \left(\Delta_{n}, d_n, A_{n}\right) \right], d, A\right)
  \]
  where $d$ and $d_i$'s indicate the modes of a construct and its arguments respectively.
  For brevity, we write $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ to indicate an operation $o$ with its arity. 
\end{definition}

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \vdash_{\Sigma, \Omega} t^d}
  \end{mathpar}

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
    \quad \text{where $d$ is either $\chk$ or $\syn$, $\Theta \vdash_{\Sigma} \Gamma$, $\Theta \vdash_{\Sigma} A$, and $\Theta \vdash_{\Sigma, \Omega} t^d$} 
     \\
    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\syn}\;(\var)
    \and
    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t : B}:  A^\syn}\;(\text{anno})
    \and
    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
    \and
  \end{mathpar}
  \begin{mathpar}
    \inferrule{\rho\colon \Sub{\Xi}{\Theta} \\
      \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
      \cdots \\
    \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
  {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{\termOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\termOp)
    \\ \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
  \end{mathpar}
  Abbreviations: $\boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$
  \caption{Bidirectional extrinsic typing rules}
\end{figure}

>>>>

\documentclass[acmsmall,screen]{acmart}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[british]{babel}

@@

\subsubsection{Algebraic theory of type theories}
\cite{Fiore1999,Fiore2022,Ahrens2021,Arkor2020,Hamana2011}

@@

More importantly, the inclusion of the set of variables used in an operation is a salient feature of our binding signatures.
Instead of treating application as a family of constructs $\app_{A, B}$ indexed by all types $A$ and $B$, as done by \citet{Fiore2022}, we are able to identify them as a single construct $\app$.
This is not only for brevity but also necessary to compare the type equality during type synthesis and checking.

@@

The substitution $\rho\colon \Sub{\Xi}{\Theta}$ is used to instantiate variables in the local context $\Xi$ with types within the context of type variables in $\Theta$.
Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\termOp$ to ensure all types are well-formed under $\Theta$.

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A} \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$ and $\Theta \vdash_{\Sigma} A$} \\
    \inferrule{A \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
    \and
    \inferrule{\rho : \Sub{\Xi}{\Theta} \\ \Theta \mid \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Theta \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\termOp)
    \and \text{for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
  \end{mathpar}
  \caption{Intrinsic typing rules}
  \label{fig:intrinsic-typing}
\end{figure}
\subsection{Raw Terms and Extrinsically Typed Terms?}
\begin{LTannote}{no extrinsic typing}
Argue that we don't need extrinsically typed terms but raw terms only as the output of some parser.
\end{LTannote}

@@

\begin{definition}
  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Cxt_{\Sigma}(\Xi) \times \Mode \times \Type_{\Sigma}(\Xi)\right)^* \times \Mode \times \Type_{\Sigma}(\Xi).
  \]
  where $\Mode$ consists of $\chk$ for checking and $\syn$ for synthesis.
  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\termOp_o$ in a bidirectional system, i.e.
  an arity is a $4$-tuple
  \[
    \left(\Xi, \left[\left(\Delta_1, d_1, A_1\right), \ldots, \left(\Delta_{n}, d_n, A_{n}\right) \right], d, A\right)
  \]
  where $d$ and $d_i$'s indicate the modes of a construct and its arguments respectively.

  For brevity, we write $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ to indicate an operation $o$ with its arity. 
\end{definition}

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}}
  \end{mathpar}

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
    \quad \text{where $d$ is either $\chk$ or $\syn$, $\Theta \vdash_{\Sigma} \Gamma$, $\Theta \vdash_{\Sigma} A$, and $\Theta \vdash_{\Sigma, \Omega} t^d$} 
    \\
    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\syn}\;(\var)
    \and
    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t : B}:  A^\syn}\;(\text{anno})
    \and
    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
  \end{mathpar}
  \begin{mathpar}
    \inferrule{\rho\colon \Sub{\Xi}{\Theta} \\
      \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
      \cdots \\
    \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
  {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{\termOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\termOp)
    \and \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
    \\
    \text{Abbreviations: $\boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$}
  \end{mathpar}
  \caption{Bidirectional extrinsic typing rules}
\end{figure}
